---
title: "ECON_104_Project_3"
author: "Marc Luzuriaga, Takuya Sugahara, Daniel Day, Shabib Alam"
date: "2023-12-08"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
#Installing the POE5data set for the first time

#install.packages("devtools")
#devtools::install_github("ccolonescu/POE5Rdata")
library (POE5Rdata)

#install.packages("AER")
library(AER)

#Other Libraries
#install.packages("xtable")
library(xtable) # makes data frame for kable, which is a package to print tables

#install.packages("jtools")
library(jtools) #for short display of the regressions 
# # This allows to use summ(Reg) instead of summary(reg)
# set_summ_defaults(digits = 3) #decimal ditits of numerical outputs

#install.packages("fitdistrplus")
library(fitdistrplus)

#install.packages("stats4")
library(stats4)

#install.packages("MASS")
library(MASS)

# for other necessary test or graphical tools
#install.packages("survival")
library(survival)

#install.packages("actuar")
library(actuar)

#install.packages("distrMod")
library(distrMod)

#install.packages("effects") # Useful package to plot marginal effects
library(effects)

#install.packages("car")
library(car)

#install.packages("knitr")
library(knitr)

library(jtools)
#install.packages("dplyr")
#install.packages('tseries')
#install.packages('forecast')
#install.packages('ggplot2')
library(tseries)
library(forecast)
library(ggplot2)
#install.packages("plm")
library(plm)
library(dplyr) 
#install.packages("gplots")
library(gplots)

#install.packages("stargazer")
library(stargazer)
```

```{r}
data(Cigar)
```

# [**1 Panel Data: Cigar**]{.underline}

## 1(a) Question

For the purposes of this project, we will ask "Which of the following
variables over time: (1) population above the age of 16, (2) consumer
price index, (3) state, (4) per capita disposable income (5) sales, (6)
per capita disposable income, or (7) minimum price in adjoining states
per pack of cigarettes have the largest impact on the price of
cigarettes over time?"

In section 2, we will express our question more precisely using the
following models: (1) Pooled Model, (2) Fixed Effects, (3) Random
Effects, and we will determine which model is best fit to answer our
question.

## 1(b) Data Set Summary

In this paper, we will be analyzing the Cigar Data Set from the PLM
Package. The data set holds 46 time observations from 1963 to 1992 and
1380 cross-sectional units about Cigarette Consumption data in the
United States. The dataset contains panel data with the following nine
variables:

(1) state: state abbreviation. For the purposes of this project and
    simplification, we will only analyze 5 states (state 5, state 33,
    state 44, state 10, and state 39) in the panel data

(2) year: the year.

(3) price: price per pack of cigarettes.

(4) pop: population

(5) pop16: population above the age of 16.

(6) cpi: consumer price index (1983=100).

(7) ndi: per capita disposable income

(8) sales: cigarette sales in packs per capita

(9) pimin: minimum price in adjoining states per pack of cigaretttes

Since we have 46 time observations (T) and 1380 cross-sectional units
(N), this panel dataset can be characterized as [**short and
wide**]{.underline}.

## 1(c) Variable Description

### 1(c).1 Graphs

The graph below illustrates a histogram displaying the frequencies with
respect to price. The graph portrays the fact that the majority of our
observations for will fall between approximately between 0 to 50 gallons
of consumption. The data also appears to right-skewed, indicating that
we would need to perform a log transformation in order to normalize the
data.

```{r}
filtered_data <- filter(Cigar, state == 5 | state == 33 | state == 44 | state == 10
                        | state == 39)
plotdist(filtered_data$price, histo = TRUE, demp = TRUE)
```

The graph below illustrates a histogram displaying the frequencies with
respect to population. The data portrays that the relationship is bi
modal (having two peaks).

```{r}
plotdist(filtered_data$pop, histo = TRUE, demp = TRUE)
```

The graph below illustrates a histogram displaying the frequencies with
respect to population over 16 years old. The data portrays that the
relationship is bi modal (having two peaks).

```{r}
plotdist(filtered_data$pop16, histo = TRUE, demp = TRUE)
```

The graph below illustrates a histogram displaying the frequencies with
respect to cpi. The data portrays that the relationship is right skewed.
A log transformation is needed to normalize the distribution.

```{r}
plotdist(filtered_data$cpi, histo = TRUE, demp = TRUE)
```

The graph below illustrates a histogram displaying the frequencies with
respect to ndi. The data portrays that the relationship is right skewed.
A log transformation is needed to normalize the distribution.

```{r}
plotdist(filtered_data$ndi, histo = TRUE, demp = TRUE)
```

The graph below illustrates a histogram displaying the frequencies with
respect to sales. The data portrays that the relationship is left
skewed.

```{r}
plotdist(filtered_data$sales, histo = TRUE, demp = TRUE)
```

The graph below illustrates a histogram displaying the frequencies with
respect to pimin. The data portrays that the relationship is right
skewed. A log transformation is needed to normalize the distribution.

```{r}
plotdist(filtered_data$pimin, histo = TRUE, demp = TRUE)
```

Finally, we present the correlation matrix. The correlation between
population and price is 0.3106. The correlation between the population
over 16 and price is 0.3742. The correlation between cpi and price is
0.9477. The correlation between ndi and price is 0.9559. The correlation
between the price and sales is -0.7448. The correlation between pimin
and price is 0.99067.

```{r}
#Correlation Matrix
my_data <- filtered_data[, c(3,4,5,6,7,8,9)]
cor(my_data)
```

### 1(c).2 Data Visualization and Heterogeneity

In this section, we will explore heterogeneity across time (year) and
states (id) as well as provide visualization of the variables against
log of price in our dataset. Below, we have a scatterplot regressing the
log of price against the year subject to the state. The graph below
presents the differences or heterogeneity of price over time among the
time-invariant variable that is states. We proceed to run the following
commands below:

```{r}
#Heterogeneity across Time
scatterplot(log(price) ~ year|state, data = filtered_data)
```

As depicted by the graph, there are differences among the intercepts of
the states across time, implying that there is heterogeneity between the
states with respect to time.

Next, we plot the means of log of price against the states. The graph
below presents the differences or heterogeneity of average price against
firms. We proceed to run the following commands below:

```{r}
#Heterogeneity across Firms
plotmeans(log(price) ~ state, data = filtered_data)
```

As depicted by the plot above, there are differences and heterogeneity
among the mean price across firms.

Next, we graph a scatter plot of the log of price against the year.

```{r}
scatterplot(log(price) ~ year, data = filtered_data)
```

As depicted by the graph above, we see that the relationship between log
of price and year has a positive linear relationship. We also see that
the median log of price is approximately 4.0.

Finally, the rest of this section plots the relationship between price
and the rest of the variables, which include "pop", "cpi", "ndi",
"sales", and "pimin":

The following is a scatterplot of log(price) vs log(pop). The
scatterplot indicates a positive but highly variable relationship
between the two variables.

```{r}
scatterplot(log(price) ~ log(pop), data = filtered_data)
```

The following is a scatterplot of log(price) vs log(pop16). The
scatterplot indicates a positive but highly variable relationship
between the two variables.

```{r}
scatterplot(log(price) ~ log(pop16), data = filtered_data)
```

The following is a scatterplot of log(price) vs log(cpi). The
scatterplot indicates a positive but tight relationship between the two
variables.

```{r}
scatterplot(log(price) ~ log(cpi), data = filtered_data)
```

The following is a scatterplot of log(price) vs log(ndi). The
scatterplot indicates a positive but tight relationship between the two
variables.

```{r}
scatterplot(log(price) ~ log(ndi), data = filtered_data)
```

The following is a scatterplot of log(price) vs log(ndi). The
scatterplot indicates a positive relationship between the two variables.

```{r}
scatterplot(log(price) ~ log(ndi), data = filtered_data)
```

The following is a scatterplot of log(price) vs log(sales). The
scatterplot indicates a positive but highly variable relationship
between the two variables.

```{r}
scatterplot(log(price) ~ log(sales), data = filtered_data)
```

The following is a scatterplot of log(price) vs log(pimin). The
scatterplot indicates a positive but tight relationship between the two
variables.

```{r}
scatterplot(log(price) ~ log(pimin), data = filtered_data)
```

# [**2 Panel Data Models: Cigar**]{.underline}

## 2(a) Pooled Model

In this section, we will be creating a pooled model to model the
question that we provided in section 1(a). We proceed to plot the pooled
regression with the following commands below:

```{r, warning=FALSE}
pooledreg1 <- plm(price~pop+pop16+cpi+ndi+sales+pimin,model="pooling",data=filtered_data)
crse<- coeftest(pooledreg1, vcov=vcovHC(pooledreg1,
type="HC0",cluster="group"))
stargazer(pooledreg1, crse, column.labels = c("\\textit{Pooled}", "\\textit{Pooled(prse)}"),
model.names = FALSE,type = "text")
```

From the pooled regression model, we could note that cpi, sales, pimin,
and the constant are statistically significant. However, pop, pop16, and
ndi are not statistically significant. We also note that the R-squared
is reasonably high with the value of 0.984 being between 0.5 and 0.99.

However, the estimated pooled model assumes that the intercept value
between the states are the same. Recall our observation from 1(c).2 that
the data exhibits heterogeneity across firms and across time. This
indicates to us that we need to relax the first assumption that the
slope coefficients are constant but the intercepts varies across the
cross sectional units. Thus, in the subsequent models, we will relax
this constraint by using the fixed effects and random effects model in
order to better model our question that we presented in the beginning of
the paper.

## 2(b) Fixed Effects Model

```{r}
fixed_effects_model <- plm(price ~ pop + pop16 + cpi + ndi + sales + pimin,model = 
                             "within", data = filtered_data)
summary(fixed_effects_model)
```

The dataset has a balanced panel with 5 entities (cross-sectional
units), 30 time periods, and a total of 150 observations.The residuals
(differences between observed and predicted values) have a minimum of
-21.19 and a maximum of 14.51. Interpretations for significant
variables: sales: For a one-unit increase in sales, the price decreases
by approximately 0.3124 units. pimin: For a one-unit increase in pimin,
the price increases by approximately 0.9951 units. sales and pimin are
highly statistically significant (p-values \< 0.01), indicating their
strong impact on the dependent variable. The R-squared value is 0.9864,
suggesting that the Fixed Effects model explains a substantial portion
of the variation in the dependent variable. The F-statistic is 1677.9
with a very low p-value, indicating that the overall model is
statistically significant. \## 2(c) Random Effects Model

```{r}
cor(filtered_data[, c("price", "pop", "pop16", "cpi", "ndi", "sales", "pimin")])
```

Here we can see that there has high multicoleanirity between price\~cpi,
price\~ndi, and price\~pimin. Need to remove those three variables that
are causing the problem.

```{r}
random_effects_model <-plm(price ~ pop + pop16 + sales,model = "random", data = filtered_data)
summary(random_effects_model)
```

The Random Effects model provides a reasonable fit to the data,
considering the significant variables and the explained variation in
'price.' The inclusion of individual effects helps capture
entity-specific variability, addressing the heterogeneity observed in
the dataset.

## 2(d) Which Model Should We Use?

1.  Pooled Model: R-Squared: 0.984 Adjusted R-Squared: 0.983
    F-Statistic: 1,464.068\*\*\* (p-value \< 0.01) Significant
    Coefficients: cpi, sales, pimin (at various significance levels)

2.  Fixed Effects Model: R-Squared: 0.98638 Adjusted R-Squared: 0.9854
    F-Statistic: 1,677.9\*\*\* (p-value \< 2.22e-16) Significant
    Coefficients: pop, pop16, sales, pimin (at various significance
    levels)

3.  Random Effects Model: R-Squared: 0.71884 Adjusted R-Squared: 0.71306
    Chisq: 373.281 (p-value \< 2.22e-16) Significant Coefficients:
    Intercept, pop, pop16, sales (at various significance levels)

Based on the results and considerations, the Fixed Effects Model seems
to be the preferred choice in this scenario. It has a higher Adjusted
R-squared, a significant F-statistic, and all coefficients are
significant.

# [**3 Conclusion: Cigar**]{.underline}

The Fixed Effects Model seems to provide the best fit (highest Adjusted
R-squared). The Pooled Model performs well but may overlook
individual-specific effects. The Random Effects Model, while capturing
individual-specific effects, has a lower overall fit compared to the
Fixed Effects Model.

Preferred Model: Fixed Effects Model High Adjusted R-squared. Captures
individual-specific effects. Addresses potential heterogeneity across
entities.

------------------------------------------------------------------------

# [**4 Binary Dependent Variables: Credit Card**]{.underline}

## 4(a) Question

```{r}
data("CreditCard")
```

For the purposes of this project, we will ask "Which of the following
variables : (1) report, (2)age, (3) income, (4)expenditure,
(5)dependents, (6) months, (7) active, have the largest impact on the
whether someone will obtain a new Credit Card?"

In section 2, we will express our question more precisely using the
following models: (1) Linear probability model, (2) Probit Model, (3)
Logit model, and we will determine which model is best fit to answer our
question.

## 4(b) Data Set Summary

In this paper, we will be analyzing the Credit Card Data Set from the
PLM Package. Credit Card data in the United States. The dataset contains
panel data with the following twelve variables:

(1) card: Factor. Was the application for a credit card accepted?

(2) reports: Number of major derogatory reports

(3) age: Age in years plus twelfths of a year

(4) income: Yearly income (in USD 10,000)

(5) share: Ratio of monthly credit card expenditure to yearly income.

(6) expenditure: Average monthly credit card expenditure

(7) owner: Factor. Does the individual own their home?

(8) selfemp: Factor. Is the individual self-employed?

(9) dependents: Number of dependents

(10) months: Months living at current address.

(11) majorcards: Number of major credit cards held

(12) active: Number of active credit accounts

## 4(c) Variable Description

### 4(c).1 Graphs

The graph below illustrates a histogram displaying the frequencies with
respect to price. The graph portrays the fact that the majority of our
observations for will fall between approximately between 0 to 50 gallons
of consumption. The data also appears to right-skewed, indicating that
we would need to perform a log transformation in order to normalize the
data.

```{r}
card <- table(CreditCard$card)
barplot(card)
CreditCard$card<-(CreditCard$card=="yes")*1
```

```{r}
#Change the variable of "Age"
plotdist(CreditCard$age, histo = TRUE, demp = TRUE)
```

The graph above illustrates a histogram displaying the frequencies with
respect to age. The data portrays that the relationship is left skewed.

```{r}
plotdist(CreditCard$income, histo = TRUE, demp = TRUE)
```

The graph above illustrates a histogram displaying the frequencies with
respect to income (USD 10,000). The data portrays that the relationship
is right skewed

```{r}
plotdist(CreditCard$expenditure, histo = TRUE, demp = TRUE)
```

The graph above illustrates a histogram displaying the frequencies with
respect to expenditure. The data portrays that the relationship is right
skewed.

```{r}
plotdist(CreditCard$dependents, histo = TRUE, demp = TRUE)
```

The graph above illustrates a histogram displaying the frequencies with
respect to dependents. The data portrays that the relationship is right
skewed.

```{r}
plotdist(CreditCard$months, histo = TRUE, demp = TRUE)
```

The graph above illustrates a histogram displaying the frequencies with
respect to months. The data portrays that the relationship is right
skewed. A log transformation is needed to normalize the distribution.

```{r}
plotdist(CreditCard$majorcards, histo = TRUE, demp = TRUE)
```

The graph above illustrates a histogram displaying the frequencies with
respect to months. The data portrays that the relationship has two peak.

```{r}
plotdist(CreditCard$active, histo = TRUE, demp = TRUE)

```

The graph above illustrates a histogram displaying the frequencies with
respect to months. The data portrays that the relationship is right
skewed.

Finally, we present the correlation matrix. The relationships between
each variable in our dataset is presented in the matrix below:

```{r}
#Correlation Matrix
data1 <- CreditCard[, c(3,4,5,6,9,10,12)]
cor(data1)
```

# [**5 Binary Dependent Variables: Credit Card**]{.underline}

## 5(a) Linear Probability Model

The Linear Dependent Variable Model is presented below:

```{r}

linear_model <- lm(card ~ reports + age + income + expenditure + dependents + months
                   + active, data = CreditCard)

summary(linear_model)

AIC(linear_model)
BIC(linear_model)
```

## 5(b) Probit model

The Probit Model is presented below:

```{r, warning=FALSE}

probit_model <- glm(card ~ reports + age + income + expenditure + dependents + 
                      months + active, family = binomial(link = "probit"), 
                    data = CreditCard)

summary(probit_model)
BIC(probit_model)
```

## 5(c) Logit Model

The Logit Model is presented below:

```{r, warning=FALSE}

logit_model <- glm(card ~ reports + age + income + expenditure + dependents + months
                   + active, family = binomial(link="logit"), data = CreditCard)

summary(logit_model)

BIC(logit_model)
```

## 5(d) Which Model Should We Use?

The probit is the better model with the lowest AIC and BIC of 131.88 and
173.3591 respectively, compared to the linear model (AIC = 934.0028, BIC
= 980.6645) and the logit Model (AIC = 132.12, BIC = 173.5977).

## [**6**]{.underline} Which Model Should We Use?

All three models had reports, dependents, and actives as statically
significant. Only the linear model had expenditure as significant.
Reports had the largest effect that is statically significant. This
makes sense since a history of delinquency would serious hinder someone
getting a credit card. Our group thought it was interesting that income
and expenditure did not have a significant role in determining whether
someone got approved or not, since a person with a high income better
able pay of their credit card statement and since a person who spends a
lot who probably have a great risk of spending too much on their credit
card. However, the other variables are better suited at predicting the
truth.

Thus, our recommended model would be
$Pr[Card =1] =\Phi[-1.54 - 1.11*reports + 0.02*age -0.05*income + 13.24*expenditure - 0.35*dependents - 0.002*months+ 0.05*active]$
